Smoothing Algorithms for Variable Rate Models

### 2012-04-19 draft3_VRsmoothing.tex
Final draft.

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}

For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) (apart from the first which is sampled from (\ref{eq:cp_cond_model})) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using \ref{eq:backward_kf_predict_start} to \ref{eq:backward_kf_predict_stop}.
      \FOR{$j=1 \dots N_P$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
    \FOR{$n=N \dots 1$}
      \STATE Initialise chain with $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
      \FOR{$m=1 \dots M$}
        \STATE Propose a new history $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n'^{(i)*}$.
        \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
      \ENDFOR
      \STATE Store final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{MCMC piecewise-deterministic variable rate particle smoother}
\label{alg:VRPS}
\end{figure}

@@

@@

The jump covariance matrices are,
%
\begin{IEEEeqnarray}{c}
Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
\end{IEEEeqnarray}

This model may be discretised at the observation times using standard methods (see e.g. \cite{Grewal2002}). Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
 y_n          &=& C \mathbf{x}_{n} + v_n
\end{IEEEeqnarray}

where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
%
\begin{IEEEeqnarray}{rCl}
 A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
 C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
                        Q_{D}             & \text{otherwise} \end{cases} \\
 Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
 q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
 q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
 q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
 R     &=& [\sigma_y^2]
\end{IEEEeqnarray}

@@

The algorithms were first tested on artificial data simulated from this model. The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.

The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter.

An example realisation simulated from the model is shown in figure~\ref{fig:example_data}.

@@

Both the smoother and the final processing step of the filter produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. It is actually no simple task to compare the quality of these approximations. Each particle consists of a list of changepoint times and types, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, once a best estimate has been calculated, it is not obvious how the error should be assessed, as it is not trivial to define a distance metric between lists of varying lengths.

For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times. These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles. The resulting function provides an approximate measure of the likelihood of finding a jump at a given time. These are shown in figure~\ref{fig:example_kdest} for the example data set.

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
\label{fig:example_kdest}
\end{figure}

It is immediately apparent that the approximation produced by the filter lacks diversity. The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between. This is a result of every particle containing an almost identical list of changepoints, as a result of resampling. In contrast, the smoother is clearly providing a less degenerate particle representation, and generally contains larger kernel density peaks at the times of larger jumps. Furthermore, the kernel density for the trend jump times contains broader peaks, capturing the fact that because the trend is not observed directly, the jumps are harder to localise precisely.

A further comparison of the algorithms is possible via the state estimates they generate. Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother. For the example data set, the trend estimates are shown in figure~\ref{fig:example_state} (The estimates of value are less informative, as this quantity is observed). Again, the improved particle diversity of the smoother is apparent.

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}

>>>>

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}. An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.  %, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

%For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
%  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
%\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula. % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using \ref{eq:backward_kf_predict_start} to \ref{eq:backward_kf_predict_stop}.
      \FOR{$j=1 \dots N_P$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
    \FOR{$n=N \dots 1$}
      \STATE Initialise chain with $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
      \FOR{$m=1 \dots M$}
        \STATE Propose a new history $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n'^{(i)*}$.
        \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
      \ENDFOR
      \STATE Store final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{MCMC piecewise-deterministic variable rate particle smoother}
\label{alg:VRPS}
\end{figure}

@@

The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. Calculating a single estimator from these is no simple task. Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, if a best estimate could be calculated, it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the root mean squared error (RMSE) of the resulting estimators.

In the following examples, we compare the state estimates from the VRPF, the ``filter-smoother'' and the VRPS. Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution. In the linear-Gaussian case, an RTS smoother is used to find the state posterior distribution for each particle, $p(x_n|\theta^{(i)}, y_{1:N})$, from which state estimates are derived.

In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, the number of unique state estimates is plotted over time in the following examples.

@@

%The jump covariance matrices are,
%%
%\begin{IEEEeqnarray}{c}
%Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
%                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
%\end{IEEEeqnarray}

This model may be discretised at the observation times using standard methods (see e.g. \cite{Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
 y_n          &=& C \mathbf{x}_{n} + v_n
\end{IEEEeqnarray}

where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
%
%\begin{IEEEeqnarray}{rCl}
% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
%\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
%                        Q_{D}             & \text{otherwise} \end{cases} \\
% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
% R     &=& [\sigma_y^2]
%\end{IEEEeqnarray}

@@

The algorithms were tested on artificial data simulated from this model. Observations were made on a regular time grid with a spacing of $0.0017s$. The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5. The diffusion constant is 0.05 and the observation noise standard deviation 0.001. The jump standard deviations are 0.005 and 0.05 respectively for value and trend.

%The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.

The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter. An example realisation simulated from the model is shown in figure~\ref{fig:example_data}. The trend estimates produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value estimates are not particularly informative, as this quantity is observed).

@@

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}

%For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times. These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles. The resulting function provides an approximate measure of the likelihood of finding a jump at a given time. These are shown in figure~\ref{fig:example_kdest} for the example data set.

%\begin{figure}[!t]
%\centering
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
%\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
%\label{fig:example_kdest}
%\end{figure}

%It is immediately apparent that the approximation produced by the filter lacks diversity. The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between. This is a result of every particle containing an almost identical list of changepoints, as a result of resampling. In contrast, the smoother is clearly providing a less degenerate particle representation, and generally contains larger kernel density peaks at the times of larger jumps. Furthermore, the kernel density for the trend jump times contains broader peaks, capturing the fact that because the trend is not observed directly, the jumps are harder to localise precisely.

%A further comparison of the algorithms is possible via the state estimates they generate. Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother. For the example data set, the trend estimates are shown in figure~\ref{fig:example_state} (The estimates of value are less informative, as this quantity is observed). Again, the improved particle diversity of the smoother is apparent.

%\begin{figure}[!t]
%\centering
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
%\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
%\label{fig:example_state}
%\end{figure}
